A Theory for Data Migration of Generated Evolving Information Systems

### 2022-10-29 2022 Migration/articleMigration.tex
redone sections to fit schemas into the picture.

   We recognize that designs from which concrete systems are generated can have errors in them:
   Either a designer was making a mistake directly,
   or the requirements the designer was taking into account were miscommunicated,
   or the environment changed in a way that was not foreseen.
   Regardless of the reason, even a generated system may need to be repaired.
   For this reason, we focus on evolving systems.

@@

   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
   A system $\infsys$ contains a dataset $\dataset_{\infsys}$, which represents the state of the system.
   (In the sequel we may just write $\dataset$
   if the corresponding system is either obvious from the context or irrelevant for the argument.)
   Every event that an actor causes in system $\infsys$ may change the state $\dataset$ of that system.
   Data migration occurs when an old system, $\infsys$, is replaced by a new system, $\infsys'$,
   while preserving the meaning of the data from $\infsys$ as much as possible.
   Just copying the set of data from $\infsys$ to $\infsys'$ is obviously wrong if the schemas of both systems differ.

@@

   Extra complications arise from accumulated data pollution in the old system.
   This allows us to categorize data pollution in the old system in the following categories:
   \begin{itemize}
      \item data pollution that cannot be captured in constraints on the dataset,
      such as a street address that has become obsolete because the person failed to send a change of address.
      \item data pollution that is a violation of a constraint in the schema.
      \item data pollution that is a violation of a constraint that is not in the schema.
      \item data pollution that is a consequence of an erroneous constraint in the schema.
   \end{itemize}
   The first category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   Such constraints move this type of pollution to one of the other categories.
   The second category, violations of constraints from the schema, does not occur in our work.
   Since the old system $\infsys$ was generated by Ampersand,
   $\infsys$ keeps all constraints from the schema satisfied at all times.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   Data pollution that is not captured by any of the constraints mentioned in the schema
   can be dealt with by adding constraints in the new system $\infsys'$.
   In that case, a migration engineer must ensure that $\infsys'$ can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary.
   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in $\infsys'$.

@@

\subsection{Datasets}
\label{sct:Datasets}
   A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.

   Before defining datasets, we must first define the constituent notions of atom, concept, triple, and relation.

   Atoms serve as data elements.
   They are values without internal structure of interest, meant to represent data elements in a database.
   From a business perspective, atoms represent concrete items of the world,
   such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
   The set of atoms is called $\atoms$.

   Concepts are names we use during the design of a dataset
   to name a group of atoms of the same type,
   to assign signatures to relations, and thus
   to enable type checking of the dataset.
   For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
   We will use variables $A$, $B$, $C$, $D$ to represent concepts.
   The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
   The set of concepts is called $\concepts$.

   We will use triples to represent data.
   This makes our theory valid for any kind of database that triples can represent,
   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
   Each triple relates two atoms to a relation.
   A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean in practice that
   the ``thing'' that \atom{Peter} refers to has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.
   In this paper, the informal meaning is interesting only for the sake of illustration.
   The set of triples is called $\triples$.
   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.

   Relations in datasets are used to store data.
   In this paper, variables $r$, $s$, and $d$ represent relations.
   Every relation $r$ contains a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{\triples}{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
\end{equation}
   Every relation has a name, a source concept, and a target concept.
   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
   % To disassemble a relation in its name, source and target, there exist three functions:
% \[\begin{array}[3]{rcl}
   % \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
   % \id{src}&:&\rels\rightarrow\concepts\\
   % \id{tgt}&:&\rels\rightarrow\concepts
% \end{array}\]
   % Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
   % \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
   % \(\src{\declare{nm}{A}{B}}=A\), and
   % \(\tgt{\declare{nm}{A}{B}}=B\).
   The pair $\pair{A}{B}$ is called the \define{signature} of the relation.
   The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
% @Bas, de volgende twee definities zitten mij nog niet lekker.
% Zou het niet $\pop{\dataset}{\fullt{A}{B}}$ en $\pop{\dataset}{\ident{A}}$ moeten zijn (omdat $A$ en $B$ niet uit $\triples$ maar uit $\concepts$ komen)?
% Alleen,...    $\dataset$ is op deze plek nog niet gedefinieerd...
   It contains all pairs that can be made of elements of $A$ and $B$:
\[\pop{\triples}{\fullt{A}{B}}\ =\ \{ \pair{a}{b}|\ a\ \inst\ A,\ b\ \inst\ B \}\]
   The term $\ident{A}$ represents the \emph{identity relation} of concept $A$:
\[\pop{\triples}{\ident{A}}\ =\ \{ \pair{a}{a}|\ a\ \inst\ A \}\]
   The set of relations is called $\rels$ and it contains $\full$, $\iden$, and all relations mentioned in $\triples$.

   We can now define the notion of dataset.
   Let $\concepts$ be a set of concepts,
   let $\rels$ be a set of relations,
   let $\atoms$ be a set of atoms,
   and let $\inst$ be a relation $\atoms\times\concepts$ that is not in $\rels$,
   which represents the instance relation between concepts and atoms:
\begin{definition}[dataset]
   \label{def:dataset}
   \item A dataset is a tuple $\la\atoms,\concepts,\inst,\rels,\triples\ra$ with
   \begin{eqnarray}
      \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
      %\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
      \forall \declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
      \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&\begin{array}[t]{@{}l}\declare{n}{A}{B}\in\rels\ \wedge\ a\ \inst\ A\ \wedge\ b\ \inst\ B\end{array}
   \end{eqnarray}
   The set of all conceivable datasets is called $\Dataset$.
\end{definition}

\subsection{Schemas}
\label{sct:Schemas}
   Rules are used to describe whether certain (assumably desirable) properties hold in an information system.
   let $\isa$ be a partial order on $\concepts$,

\begin{definition}[rule]
   \label{def:rule}
   \item A rule $u$ is a tuple $\la \id{sign}_u:\Pair{\Concepts}{\Concepts},\id{viol}_u:\Dataset\rightarrow(\Pair{\Concepts}{\Concepts})\ra$.
   %in which $p_u$ is a constraint (i.e. a condition that constrains the triples of a dataset),
   $\id{sign}_u$ is the signature of $u$ and
   $\id{viol}_u$ is a function that gives the violations (which are pairs) of this rule,
   in which
\begin{eqnarray}
   \begin{array}[t]{rl}
      &\dataset\ =\ \la\atoms,\concepts,\inst,\rels,\triples\ra\ \wedge\\
      &\id{sign}_u=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}\\
      \Rightarrow\\
      &A\in\concepts\wedge B\in\concepts\wedge a\ \inst\ A\wedge b\ \inst\ B
   \end{array}
\end{eqnarray}
\end{definition}

   The function $\id{viol}_u$ computes the violations of $u$,
   which allows us to check at any moment in time whether the current population is free of violations.
   We say that a rule $u$ is \define{satisfied} in a dataset $\dataset$ iff $\viol{u}{\dataset}=\emptyset$.
   This allows us to define a predicate $p_u$ that indicates if a rule $u$ is satisfied:
\begin{eqnarray}
   p_u(\dataset)&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
\end{eqnarray}
   Since $\triples$ changes continually in the dataset of an information system, the violations are changing along.
   As a result, rules that are violated at one moment may be satisfied at another moment.
   The set of triples changes due to events observed by the system.
   These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.

   Every rule is assigned to a role.
   That role must ensure to satisfy that rule after some event has caused violations by changing the set of triples.
   In other words: If anyone edits in the dataset and breaks a rule, someone or something has to fix the dataset to satisfy that rule again.
   Example: if a rule says that every request must be acknowledged, the receipt of a request will break this rule.
   Sending an acknowledgement message will cause the rule to be satisfied again.

   Ampersand supports a special kind of rule that is used for specialization.
   The rule $A\isa B = \la A, A, viol_{A\isa B} \ra$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
   Instances of $A$ that are not an instance of $B$ are the violations of this rule: $viol_{A\isa B}(\dataset) = \{(a,a) \mid a\ \inst\ A \wedge \lnot (a\ \inst\ B)\}$.
   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.

   Specialization rules are special for two reasons:
   first, $A\isa B$ is automatically maintained by the system itself, by changing $\inst$ to include any violating pairs in $B$.
   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $p_{A\isa B}$ is False.
   This allows for certain run-time optimizations, which are beyond the scope of this paper.
   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
   To ensure that this does not result in problems at run-time, users are required to write specialization rules that make these comparisons possible.
   The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.

\subsection{Information Systems}
\label{sct:Information Systems}
   The purpose of an information system is to make data meaningful to users.
   Data migration is all about preserving that meaning.
   This section introduces the notion of information system
   as a system that consist of a dataset and a set of rules that constrain that data,
   together with a set of roles that must keep all rules satisfied.
   %These rules reflect the meaning (semantics) that is shared among users.

@@

   We can now define the notion of information system.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\roles,\rules,\specrules,\maintain,\dataset\ra$.
   Here $\roles$ is a set of roles,
   $\rules$ a set of rules,
   $\specrules$ a set of specialization rules,
   $\maintain$ is a relation $\roles\times\rules$,
   and $\dataset$ is a dataset.
\end{definition}
   The relation $\maintain$ links rules to roles.
   $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\triples}$ empty.
   If a rule $u$ has violations, i.e. $\viol{u}{\triples}$ is not empty, there is a role whose job it is to remove those violations.
   The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.

@@

   Now let us define an information system.
   We define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
   Its constraint, $p_{\tt Enroll}$ is written in logic as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}\]
   In Ampersand, which is a syntactically sugared form of relation algebra,
   we give each rule a name and declare a role to maintain it:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
   The Ampersand compiler derives the signature of every rule.
   This rule has signature $\pair{\text{\tt Student}}{\text{\tt Module}}$.
   The Ampersand compiler also derives the violation set as:
\[\begin{array}{l}
   \viol{\text{\tt EnrollRule}}{\triples}\\
   \hspace{1cm}=\{\pair{s}{m}|s\ \text{\tt isEnrolledFor}\ m\ \wedge\not\exists c: s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
\end{array}\]
   Notice that the current population yields an empty set of violations,
   so the rule is satisfied.
   Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
   we have now defined a small system.

@@

   We use two datasets: $\dataset$ and $\dataset'$.
   Before doing so, let us first define the disjoint union of two systems.
\begin{definition}[disjoint union of datasets]
\begin{eqnarray}
   \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\triples'\ra$}\notag\\
   &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \triples\uplus^5\triples'\ra\notag\\
      X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\
      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))|\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))|\ (y_1,y_2)\in Y\}\end{array}\\
      X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
      X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
\end{eqnarray}
\end{definition}
Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.

@@

\subsection{Changes}
% The purpose of this section is to explain why a dataset is structured the way it is.
   In the migration of a dataset we deal with changes to the elements that are not data:
   $\inst$, $\concepts$, and $\rels$.
   Such changes have further reaching consequences, however.
   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\triples\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
   To satisfy definition~\ref{def:dataset}, $\triples$ and $\atoms$ must change too,
   but the migration should try to preserve the maximal amount of data.

>>>>

   We recognize that designs from which concrete systems are generated can have errors in them:
   Either an engineer was making a mistake directly,
   or the requirements the engineer was taking into account were miscommunicated,
   or the environment changed in a way that was not foreseen.
   Regardless of the reason, even a generated system may need to be repaired.
   For this reason, we focus on evolving systems.

@@

   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
   A system $\infsys$ contains a dataset $\dataset_{\infsys}$, which represents the state of the system.
   (In the sequel we may just write $\dataset$
   if the corresponding system is either clear from the context or irrelevant for the argument.)
   Every event that an actor causes in system $\infsys$ may change the state $\dataset$ of that system.
   Data migration occurs when an old system, $\infsys$, is replaced by a new system, $\infsys'$,
   while preserving the meaning of the data from $\infsys$ as much as possible.
   Just copying the set of data from $\infsys$ to $\infsys'$ is obviously wrong if the schemas of both systems differ.

@@

   Extra complications arise from accumulated data pollution in the old system.
   This allows us to categorize data pollution in the old system in the following categories:
   \begin{itemize}
      \item data pollution that cannot be captured in constraints on the dataset,
      such as a street address that has become obsolete because the person failed to send a change of address.
      \item data pollution that violates a constraint in the schema.
      \item data pollution that violates a constraint that is not in the schema.
      \item data pollution that is a consequence of an erroneous constraint in the schema.
   \end{itemize}
   The first category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   Such constraints move this type of pollution to one of the other categories.
   The second category, violations of constraints from the schema, does not occur in our work.
   Since the old system $\infsys$ was generated by Ampersand,
   $\infsys$ keeps all constraints from the schema satisfied at all times.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   Data pollution that is not captured by any of the constraints mentioned in the schema
   can be dealt with by adding constraints in the new system $\infsys'$.
   In that case, a migration engineer must ensure that $\infsys'$ can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary.
   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in $\infsys'$.

@@

\subsection{Datasets}
\label{sct:Datasets}
   A dataset $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
   Before defining datasets, we must first define the constituent notions of atom, concept, relation, and triple.

   Atoms serve as data elements.
   They are values without internal structure of interest, meant to represent data elements in a database.
   From a business perspective, atoms represent concrete items of the world,
   such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
   All atoms are taken from an infinite set called $\Atoms$.

   Concepts are names that group atoms of the same type.
   For example, an engineer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
   and \atom{074238991} as a \concept{TelephoneNumber}.
   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
   We will use variables $A$, $B$, $C$, $D$ to represent concepts.
   The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
   All concepts are taken from an infinite set $\Concepts$.
   $\Concepts$ and $\Atoms$ are disjoint sets.

   Relations serve to organize and store data, to allow an engineer to represent facts.
   In this paper, variables $r$, $s$, and $d$ represent relations.
   All relations are taken from an infinite set $\Rels$.
   $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.

   Triples serve to represent data.
   A dataset $\dataset$ is a set of triples.
   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
   This makes our theory valid for any kind of database that triples can represent,
   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
   Each triple relates two atoms to a relation.
   A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean in practice that
   the ``thing'' that \atom{Peter} refers to has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.
   In this paper, the informal meaning is interesting only for the sake of illustration.
   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\dataset$
   in those cases where $\dataset$ is either clear from the context or irrelevant for the argument.
   A relation $r$ can serve as a container of pairs,
   as defined by the funciton $\id{pop}_r:\Dataset\rightarrow\Pair{\Atoms}{\Atoms}$.
   It defines a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\dataset\}
\end{equation}

   Over time, an information system ``observes'' events that change the dataset by inserting and deleting triples in $\dataset$.
   By the same account, the population of relations change over time.
   These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.

\subsection{Schemas}
\label{sct:Schemas}
   To design an information system, an engineer defines a schema in which she defines concepts, relations, and rules.
   Rules serve the purpose of invariants,
   i.e.\ properties that the data in the database must satisfy at all times.
   In this paper, variables $u$ and $v$ represent rules.

   A schema $\schema$ is a triple $\triple{\concepts_{\schema}}{\rels_{\schema}}{\rules_{\schema}}$,
   in which $\concepts_{\schema}$ is a finite set of concepts,
   $\rels_{\schema}$ is a finite set of relations,
   and $\rules_{\schema}$ is a finite set of rules.
   We will simply write $\concepts$, $\rels$, or $\rules$ if the schema is clear from the context or irrelevant for the argument.
   A rule in a schema serves to constrain the dataset,
   to ensure the semantic integrity of the dataset.
   For every rule $u$ in the schema, the Ampersand compiler generates code for the predicate $\id{sat}_u\{\Dataset\}$
   to determine whether rule $u$ is satisfied in dataset $\dataset$.
   The Ampersand compiler also generates code for a function $\id{viol}_u : \Dataset\rightarrow\id{Text}$
   to produce meaningful error messages in case rule $u$ is not satisfied.

\subsection{Information Systems}
\label{sct:Information Systems}
   The purpose of an information system is to make data meaningful to its users.
   This section introduces the notion of information system as a quintuple $\la\dataset_{\infsys},\schema_{\infsys},\inst_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
   As before, we will omit the suffixes if $\infsys$ is obvious (i.e. clear from the context or irrelevant for the argument).
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\inst,\roles,\maintain\ra$, in which
\begin{itemize}
   \item dataset $\dataset$ and $\schema$ are defined as in sections~\ref{sct:Schemas} and~\ref{sct:Datasets};
   \item $\roles$ is a set of roles;
   \item $\inst : \Atoms\times\Concepts$ is the instance relation between atoms from $\dataset$ and concepts from $\schema$;
   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$; and
   \item requirements~\ref{eqn:inst} thru~\ref{eqn:specialization} are satisfied.
\end{itemize}
\end{definition}
   Let us discuss these items one by one.

   The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
   This means that every atom is an instance of a concept:
\begin{eqnarray}
   \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:inst}
\end{eqnarray}
   To do type checking based on the schema only,
   all relations in triples from the dataset must be ``known'' in the schema:
\begin{equation}
   \forall a,r,b:\ \triple{a}{r}{b}\in\dataset\ \Rightarrow\ r\in\rels
\end{equation}

   In a typed information system, atoms in a relation have a type as declared in a relation.
   For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
\begin{eqnarray}
   \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
\end{eqnarray}

   To compare atoms for equality,
   the set $\rels$ must contain an identity relation $\ident{C}$ for every concept $C$ in which all atoms occur.
   It must satisfy:
\begin{equation}
   \forall a,b:\ a\ \inst\ C\ \wedge\ b\ \inst\ C\ \Rightarrow\ (a\ \ident{C}\ b\ \Leftrightarrow\ a=b)
\end{equation}
   So, $\id{pop}_{\ident{C}}$ is defined by:
\begin{eqnarray}
   \forall C\in\concepts&:&\pop{\ident{C}}{\dataset}=\{\pair{a}{a}| a\ \inst\ C\}
\end{eqnarray}

   Ampersand supports a special kind of rule that is used for specialization.
   The rule $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
\begin{equation}
   \label{eqn:specialization}
   \id{sat}_{A\isa B}\{\dataset\}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
\end{equation}
   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.

   All specialization rules in a schema together form a partial order $\isa : \Pair{\concepts}{\concepts}$.
   Specialization rules are special for two reasons:
   Firstly, the system keeps $A\isa B$ satisfied automatically by changing $\inst$ whenever needed.
   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $\id{sat}_{A\isa B}$ is False.
   This allows for certain run-time optimizations, which are beyond the scope of this paper.
   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
   To ensure that this does not result in problems at run-time, users are required to write specialization rules that make these comparisons possible.
   The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.

   %These rules reflect the meaning (semantics) that is shared among users.

   Every rule is assigned to a role.
   That role must ensure to satisfy that rule, after some event has violated the rule by changing the set of triples.
   In other words: If anyone edits in the dataset and violates a rule, someone or something has to fix the dataset to satisfy that rule again.
   Example: if a rule says that every request must be acknowledged, the receipt of a request will violate this rule.
   Sending an acknowledgement message will cause the rule to be satisfied again.

@@

   The relation $\maintain$ links rules to roles.
   $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\dataset}$ empty.
   If a rule $u$ has violations, i.e. $\viol{u}{\dataset}$ is not empty, there is a role whose job it is to remove those violations.
   The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.

@@

   Now let us define an information system.
   We define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
   Its constraint, $p_{\tt Enroll}$ is written in logic as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}\]
   In Ampersand, which is a syntactically sugared form of relation algebra,
   we give each rule a name and declare a role to maintain it:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
   The Ampersand compiler derives the signature of every rule.
   This rule has signature $\pair{\text{\tt Student}}{\text{\tt Module}}$.
   The Ampersand compiler also derives the violation set as:
\[\begin{array}{l}
   \viol{\text{\tt EnrollRule}}{\dataset}\\
   \hspace{1cm}=\{\pair{s}{m}|s\ \text{\tt isEnrolledFor}\ m\ \wedge\not\exists c: s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
\end{array}\]
   Notice that the current population yields an empty set of violations,
   so the rule is satisfied.
   Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
   we have now defined a small system.

@@

   We use two datasets: $\dataset$ and $\dataset'$.
   Before doing so, let us first define the disjoint union of two systems.
\begin{definition}[disjoint union of datasets]
\begin{eqnarray}
   \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
   &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
      X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\
      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))|\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))|\ (y_1,y_2)\in Y\}\end{array}\\
      X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
      X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
\end{eqnarray}
\end{definition}
Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.

@@

\subsection{Changes}
% The purpose of this section is to explain why a dataset is structured the way it is.
   In the migration of a dataset we deal with changes to the elements that are not data:
   $\inst$, $\concepts$, and $\rels$.
   Such changes have further reaching consequences, however.
   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
   To satisfy definition~\ref{def:dataset}, $\dataset$ and $\atoms$ must change too,
   but the migration should try to preserve the maximal amount of data.
